---
title: "Check On Yeli: May 17, 2022"
output: html_notebook
---

```{r Setup, include = F}

library(baseballr)
library(ggplot2)
library(dplyr)
library(stringr)
library(lubridate)
library(devtools)

#build a vector of MLB regular season game dates
source_url("https://raw.githubusercontent.com/JanLMoffett/baseballAnalysis/master/buildDatesMLB.R")

#a randomly sampled set of BIP to derive "average" comparison
av <- read.csv(url("https://raw.githubusercontent.com/JanLMoffett/baseballAnalysis/master/data/BIP_35dates_2015_21.csv"))

#importing colors and themes for plots
#github.com/JanLMoffett/datavizExtras
source_url("https://raw.githubusercontent.com/JanLMoffett/datavizExtras/master/colorConstants.R")
source_url("https://raw.githubusercontent.com/JanLMoffett/datavizExtras/master/extraThemes.R")
source_url("https://raw.githubusercontent.com/JanLMoffett/datavizExtras/master/datavizExtras.R")
```


```{r data1, include = F}
#looking up Christian's MLB ID (592885)
#cyID <- playerid_lookup(last_name = "Yelich", first_name = "Christian")$mlbam_id
cyID <- 592885

#getting the data from baseball savant data with baseballr
#the 2022 regular season started on Apr 7
opening_day2022 <- "2022-04-07"
cy <- statcast_search(start_date = opening_day2022, 
                end_date = "2022-05-16",
                player_type = "batter",
                playerid = cyID)

#2021
cy21 <- statcast_search(start_date = SCseasons$start[SCseasons$season == 2021],
                        end_date = "2021-06-17",
                        player_type = "batter",
                        playerid = cyID)

g35.21 <- cy21 %>% group_by(game_date) %>% 
  summarize(games = 1) %>%
  mutate(cum_games = cumsum(games)) %>%
  filter(cum_games == 35) %>%
  pull(game_date)

cy21 <- statcast_search(start_date = SCseasons$start[SCseasons$season == 2021],
                        end_date = g35.21,
                        player_type = "batter",
                        playerid = cyID)

#2019
cy19 <- statcast_search(start_date = SCseasons$start[SCseasons$season == 2019],
                        end_date = "2019-05-17",
                        player_type = "batter",
                        playerid = cyID)

g35.19 <- cy19 %>% group_by(game_date) %>% 
  summarize(games = 1) %>%
  mutate(cum_games = cumsum(games)) %>%
  filter(cum_games == 35) %>%
  pull(game_date)

cy19 <- statcast_search(start_date = SCseasons$start[SCseasons$season == 2019],
                        end_date = g35.19,
                        player_type = "batter",
                        playerid = cyID)


#2018
cy18 <- statcast_search(start_date = SCseasons$start[SCseasons$season == 2018],
                        end_date = "2018-06-17",
                        player_type = "batter",
                        playerid = cyID)

g35.18 <- cy18 %>% group_by(game_date) %>% 
  summarize(games = 1) %>%
  mutate(cum_games = cumsum(games)) %>%
  filter(cum_games == 35) %>%
  pull(game_date)

cy18 <- statcast_search(start_date = SCseasons$start[SCseasons$season == 2018],
                        end_date = g35.18,
                        player_type = "batter",
                        playerid = cyID)


#combine sets and create season variable
cy <- cy %>% union(cy21) %>% union(cy19) %>% union(cy18) %>%
  mutate(season = as.numeric(str_sub(game_date, start = 1, end = 4)))


hitEvents <- c("single","double","triple","home_run")
outEvents <- c("field_out", "fielders_choice_out", "force_out",
               "grounded_into_double_play", "double_play")

#making some variables for the summary
cy <- cy %>% 
  mutate(
    paID = paste0(game_date,"_", at_bat_number),
    
    ball = ifelse(type == "B", 1, 0),
    strike = ifelse(type == "S", 1, 0),
    callStrike = ifelse(description == "called_strike", 1, 0),
    swingStrike = ifelse(description == "swinging_strike", 1, 0),
    
    BB = ifelse(events == "walk", 1, 0),
    SO = ifelse(events == "strikeout", 1, 0),
    
    BIP = ifelse(type == "X", 1, 0),
    hit = ifelse(events %in% hitEvents, 1, 0),
    single = ifelse(events == "single",1,0),
    double = ifelse(events == "double",1,0),
    triple = ifelse(events == "triple",1,0),
    HR = ifelse(events == "home_run",1,0))
```

Christian's numbers from his first 35 games of each season:

```{r bySeason, echo = F}
#display stats by season
res1 <- cy %>% group_by(season) %>%
  summarize(
    games = n_distinct(game_date),
    pitches = n(),
    PA = n_distinct(paID),
    
    balls = sum(ball, na.rm = T),
    strikes = sum(strike, na.rm = T),
    callStrikes = sum(callStrike, na.rm = T),
    swingStrikes = sum(swingStrike, na.rm = T),
    
    BB = sum(BB, na.rm = T),
    SO = sum(SO, na.rm = T),
    
    BIP = sum(BIP, na.rm = T),
    hits = sum(hit, na.rm = T),
    singles = sum(single, na.rm = T),
    doubles = sum(double, na.rm = T),
    triples = sum(triple, na.rm = T),
    HR = sum(HR, na.rm = T)
    ) %>%
  mutate(
    BABIP = hits/BIP,
    bbRt = BB/PA,
    soRt = SO/PA
    
  ) %>% arrange(desc(season))

res1
    
```



Launch characteristics of Christian's balls in play, after the first 35 games of each season:

```{r BIPstats, eval = T, echo = F, results = "show"}

#BIP
res2 <- cy %>% filter(type == "X") %>%
  group_by(season) %>%
  summarize(
    BIP = n(),
    avgLaunchAngle = mean(launch_angle, na.rm = T),
    avgLaunchSpeed = mean(launch_speed, na.rm = T)) %>% 
  arrange(desc(season))

res2

```

Christian Yelich hits the ball extremely hard.  Launch speed has separated him from the pack throughout his career and continues to do so.  His contact with baseballs remained consistently punishing, even during his least productive periods.  In a separate analysis I concluded that the main culprit for his drop in production was refusing to swing, and I have a theory as to why, which I'll discuss later.  

Of course average launch angle was also at its highest during Beast Year, 2019, rising at last to the coveted fliner profile in the teens of degrees.  This lift was the final ingredient his power needed to explode.  Hard-hit balls will inevitably travel further when the ground isn't in their path.  When Christian hits balls in the air, as hard as he hits them, they elude gloves and leave parks.      

Even with this fact, I understand Christian when he says he doesn't want to hear about launch angle.  In a linear equation, raising his launch angle raises his wOBA, his home run totals, and his spirits without complication.  In reality, in the muscle and bone and reflexes of a hitter's physiognomy, raising launch angle--if you can even do it on purpose--means changing a swing, which means yanking blindly on a string with a million unseen attachments.  

Becoming a Major League Baseball player is virtually impossible.  Even for good baseball players at the highest levels of amateur, minor league, independent or international play, the odds of making it to and remaining part of The Show are dauntingly long.  The way a Major League hitter moves his body when he swings is the result of a natural selection process, the precise choreography that was necessary to get him to that point.  He does things the way he does them because all the other ways he tried along his lifelong honing to MLB form didn't produce, or weren't sustainable.  Go back in time and alter any Major League hitter's swing as it coalesces, and I bet they're not a Major Leaguer in the new, darker timeline you create.  Their swing is their swing because it's the swing that got and keeps them there.  Any variation is a gamble, and possibly a disaster, no matter how many quantitative analysts advocate for it.

In the case of Christian Yelich, The lowering of launch angle is more of a return to roots than a need for concern.  I see the hard ground balls he's been hitting and think "Classic Yeli!"  He seems to have come back to center to regroup, relaxing back into the mechanics that come most naturally to him, and it's serving him well.     


Launch characteristics of Christian's hits:

```{r hitStats, eval = T, echo = F, results = "show"}

#averages from samplingMLB file
w <- data.frame(
  type = c("BIP", "hit", "single", "double", "triple", "HR"),
  total = c(48527, 15797, 10136, 3177, 298, 2186),
  avgLaunchAngle = c(11.833299, 11.620858,  6.128548, 16.581866, 19.448190, 28.350927),
  avgLaunchSpeed = c(87.90092, 93.47043, 89.87291, 97.43947, 96.87187, 103.62863))  

#hits
cy %>% filter(events %in% hitEvents) %>%
  group_by(events) %>%
  summarize(
    count = n(),
    avgLaunchAngle = mean(launch_angle, na.rm = T),
    avgLaunchSpeed = mean(launch_speed, na.rm = T))

```

Compare to Statcast-era averages for hits:

```{r hitAvgs, eval = T, echo = F, results = "show"}

#averages from samplingMLB file
w
  
```

Game by game stats:

```{r byGame, echo = F}
#calculating stats by game
byGame <- cy %>% 
  group_by(game_date) %>% 
  summarize(
    pitches = n(),
    PA = n_distinct(paID),
    
    balls = sum(ball, na.rm = T),
    strikes = sum(strike, na.rm = T),
    callStrikes = sum(callStrike, na.rm = T),
    swingStrikes = sum(swingStrike, na.rm = T),
    
    BB = sum(BB, na.rm = T),
    SO = sum(SO, na.rm = T),
    
    BIP = sum(BIP, na.rm = T),
    hits = sum(hit, na.rm = T),
    singles = sum(single, na.rm = T),
    doubles = sum(double, na.rm = T),
    triples = sum(triple, na.rm = T),
    HR = sum(HR, na.rm = T)
             ) %>%
  mutate(
    game_seq = seq_along(unique(cy$game_date)),
    
    cumPitches = cumsum(pitches),
    cumPA = cumsum(PA),
    
    cumBB = cumsum(BB),
    cumSO = cumsum(SO),
    
    cumBIP = cumsum(BIP),
    cumHit = cumsum(hits)
    
  )

#game by game stats on his balls in play
byGameBIP <- cy %>% 
  filter(type == "X") %>%
  mutate(
    BIPover93 = ifelse(launch_speed > 93, 1, 0),
    BIPover96 = ifelse(launch_speed > 96, 1, 0),
    BIPover99 = ifelse(launch_speed > 99, 1, 0),

    GBover96 = ifelse(launch_speed > 96 & launch_angle < 0, 1, 0),
    GBover99 = ifelse(launch_speed > 99 & launch_angle < 0, 1, 0)
        ) %>%
  group_by(game_date) %>%
  summarize(
    avgLA = mean(launch_angle, na.rm = T),
    minLA = min(launch_angle, na.rm = T),
    maxLA = max(launch_angle, na.rm = T),
    
    avgLS = mean(launch_speed, na.rm = T),
    minLS = min(launch_speed, na.rm = T),
    maxLS = max(launch_speed, na.rm = T),
    
    BIPover93 = sum(BIPover93, na.rm = T),
    BIPover96 = sum(BIPover96, na.rm = T),
    BIPover99 = sum(BIPover99, na.rm = T),
    
    GBover96 = sum(GBover96, na.rm = T),
    GBover99 = sum(GBover99, na.rm = T)
            )
         
#adding BIP to other game stats                               
byGame <- byGame %>% left_join(byGameBIP, by = "game_date")
  
byGame
```

## Plots

```{r byGamePlotSetup, include = F}

#pitch by pitch BIP data
p1 <- ggplot(cy %>% filter(type == "X")) + digiturf
cap1 <- "Source: Baseball Savant"
subt1 <- "Christian Yelich's Balls in Play, 2022 Season"
  
#color scale vectors
v <- c(jmbn["periwinkle"], jmbn["highlighter"])
names(v) <- c("0","1")
v2 <- c(jmbn["navy"], transpa(jmbn["highlighter"], 100))
names(v2) <- c("0","1")

```

```{r ScatterSpeedByAngle, echo = F}

#launch speed by launch angle
p1 + geom_point(aes(x = launch_angle, y = launch_speed, color = factor(hit))) +
  scale_color_manual(values = v) + 
  labs(title = "Launch Speed by Launch Angle",
       subtitle = subt1, caption = cap1)

```

```{r BoxAngle, echo = F}

#boxplot of launch angle
p1 + geom_boxplot(aes(x = launch_angle, y = factor(hit), color = factor(hit), fill = factor(hit)), 
                  outlier.size = 3, size = 1) + 
  scale_color_manual(values = v) + 
  scale_fill_manual(values = v2) +
  labs(title = "Launch Angle", subtitle = subt1, caption = cap1)

```

```{r boxSpeed, echo = F}
#boxplot of launch speed
p1 + geom_boxplot(aes(x = launch_speed, y = factor(hit), color = factor(hit), fill = factor(hit)), 
                  outlier.size = 3, size = 1) + 
  scale_color_manual(values = v) + 
  scale_fill_manual(values = v2) +
  labs(title = "Launch Speed",
       subtitle = subt1, caption = cap1)

```

```{r histAngle, echo = F}

#histogram of launch angle
p1 + geom_histogram(aes(x = launch_angle), binwidth = 10, 
                    fill = jmbn["mauve"], color = jmbn["rose"]) + 
  labs(title = "Launch Angle",
       subtitle = subt1, caption = cap1)

#histogram of launch speed
p1 + geom_histogram(aes(x = launch_speed), binwidth = 10,
                    fill = jmbn["mauve"], color = jmbn["rose"]) + 
  labs(title = "Launch Speed",
       subtitle = subt1, caption = cap1)

```

```{r eval=F}
#game sequence time series plots
p2 <- ggplot(byGame) + digiturf

p2 + geom_line(aes(x = game_seq, y = cumPA), color = jmbn["thistle"]) + 
  geom_line(aes(x = game_seq, y = cumBIP), color = jmbn["periwinkle"])

nGames = dim(byGame)[1]
#rates
cy %>%
  summarize(
    PA = n_distinct(paID),
    BIP = sum(BIP, na.rm = T),
    hits = sum(hit, na.rm = T)
  ) %>%
  mutate(
    PAperGame = PA/nGames,
    BIPperGame = BIP/nGames,
    HperGame = hits/nGames,
    PofBIP = BIP/PA,
    PofH = hits/PA
  )


#what is he hitting and what is he missing?

#total he's seen of each pitch type
byPT <- cy %>% group_by(pitch_name) %>% summarize(totalSeen = n())

#BIP
byPTbip <- cy %>% filter(type == "X") %>%
  group_by(pitch_name) %>%
  summarize(totalBIP = n(),
            avgLaunchSpeed = mean(launch_speed, na.rm = T),
            minLaunchSpeed = min(launch_speed, na.rm = T),
            maxLaunchSpeed = max(launch_speed, na.rm = T),
            avgLaunchAngle = mean(launch_angle, na.rm = T),
            minLaunchAngle = min(launch_angle, na.rm = T),
            maxLaunchAngle = max(launch_angle, na.rm = T))

byPT %>% left_join(byPTbip, by = "pitch_name")

#hits
cy %>% filter(events %in% hitEvents) %>% 
  group_by(pitch_name) %>% tally()


#whiffs
cy %>% filter(description == "swinging_strike") %>%
  group_by(pitch_name) %>% tally()


#looking
cy %>% filter(description == "called_strike")


#by pitch location zone

cy <- cy %>% mutate(szRegion = case_when(
  
  zone %in% c(1:3) ~ "top",
  zone %in% c(4:6) ~ "mid",
  zone %in% c(7:9) ~ "bot",
  zone %in% c(11:14) ~"out",
  TRUE ~ "unk"
  
))

cy %>% select(szRegion)

byReg <- cy %>% group_by(szRegion) %>% summarize(totalSeen = n())

cy %>% filter(type == "X") %>% group_by(szRegion) %>%
  summarize(totalBIP = n(),
            avgLA = mean(launch_angle, na.rm = T),
            avgLS = mean(launch_speed, na.rm = T))


#i'm curious about how many bad calls Christian has suffered through

#strike zone boundaries
szt <- mean(cy$sz_top, na.rm = T)
szb <- mean(cy$sz_bot, na.rm = T)

szr <- 19.94/24
szl <- -19.94/24
#this is the width of the strike zone in inches, converted to feet 
#and divided by two to get dist from center

#coord to draw home plate on plot of pitch locations
home_plate <- data.frame("x" = c(-17/24, 17/24, 17/24, 0, -17/24), 
                         "y" = c(0, 0, -0.2, -0.4, -0.2))

#determining which called strikes were bad calls
#how many inches of leeway should be give the umps? 
leeway <- 1/12 #1 inches
cy <- cy %>% 
  mutate(
    szOutside = ifelse(plate_x < (szl - leeway), 1, 0),
    szInside = ifelse(plate_x > (szr + leeway), 1, 0),
    
    szLow = ifelse(plate_z < (szb - leeway), 1, 0),
    szHigh = ifelse(plate_z > (szt + leeway), 1, 0)
    
    ) %>%
  mutate(
    szOZone = ifelse(szOutside + szInside + szHigh + szLow > 0, 1, 0)
  ) %>%
  mutate(
    szBadCall = ifelse(szOZone + callStrike > 1, 1 , 0)
  )

#color scale vector
cv1 <- c(jmbn["rose"], jmbn["highlighter"])
names(cv1) <- c("0", "1")

#plot of pitch locations, called strikes with bad calls highlighted
ggplot(data = cy %>% filter(description == "called_strike")) + 
  
  coord_cartesian(xlim = c(-4,4), ylim = c(-1,7)) + 
  digiturf + 
  
  geom_vline(xintercept = c(szl, szr), color = jmbn["periwinkle"], size = 0.5, linetype = 2) +
  geom_hline(yintercept = c(szt, szb), color = jmbn["periwinkle"], size = 0.5, linetype = 2) +
  annotate("rect", xmin = szl, xmax = szr, ymin = szb, ymax = szt, fill = transpa(jmbn["periwinkle"], 80)) +
  
  annotate("text", x = -3.9, y = -0.2, label = "Catcher's Perspective", color = jmbn["blush"], hjust = 0) +
  
  annotate("rect", xmin = -5, xmax = 5, ymin = -2, ymax = 0, fill = transpa(jmbn["mauve"], 100)) + 
  annotate("polygon", x = home_plate$x, y = home_plate$y, fill = transpa(jmbn["blush"], 100), color = jmbn["mauve"]) + 
  
  labs(title = "Locations of Called Strikes",
       subtitle = "Christian Yelich, 2022 Season",
       x = "Horizontal Location (ft)", 
       y = "Vertical Location (ft)") +
  
  geom_point(aes(x = plate_x, y = plate_z, color = factor(szBadCall)), shape = 16) + 
  scale_color_manual(values = cv1)
  
#looking closer at these bad calls, and what effect they had on the plate appearance
badCallpaID <- cy %>% filter(szBadCall == 1) %>% pull(paID)

badCallPA <- cy %>% filter(paID %in% badCallpaID)
#6 out of 8 of these ended in strikeouts, the other 2 were walks

badCallPA %>% arrange(paID, pitch_number) %>% 
  select(paID, pitch_number, description, szBadCall, szLow, szOutside, events)


```
